\begin{center}\textbf{\small Acknowledgements}\end{center}

First and foremost, I would like to thank Dr. Kristijan Armeni.
His consistent dedicated assistance and lightning-fast responses allowed this project to come to fruition.
The lively discussions we were not only valuable for the thesis, but also provided me with important insights into academia as well.
Moreover, I am very thankful for the continued support in reading and improving almost every section of this thesis.
Thank you for the mentorship, support and understanding.

Furthermore, I wish to express sincere gratitude to Prof. Dr. Christopher Honey.
I am incredibly thankful for Prof. Honey's willingness to supervise this thesis.
His invaluable feedback had an immense impact on the thesis.
Without Prof. Honey's assistance and encouragement it would not have been possible to complete the thesis.
Thank you for the feedback, ideas, comments and the mentorship.

I would like to express my sincere thanks to Fabio Frohberg, Marc Gröling and Friedemann Lohss for their constant support in times of need.
Their willingness to read over the thesis again and again has proven to be a huge alleviation during the final steps of submission.
Thank you for your support and friendship.

Finally, I would like to thank Dr. Tyler Tomita and Savannah Born for reviewing and suggesting synonyms for the wikitext-dataset, and Prof. Dr. Honey and Dr. Armeni for reviewing and suggesting synonyms for the nonce-dataset.

This document is based on the thesis-template created by the working group Software Engineering at Freie Universität Berlin. It can be accessed at \url{https://git.imp.fu-berlin.de/agse/thesis-template/}.
